Telemetry Module for DTU Roadrunners' Solar Car

### 2021-06-20 documentation/FinalPaper/paper.tex

\usepackage[pdftex]{graphicx}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{float}
\usepackage{comment}
\usepackage{textcomp}
\usepackage{siunitx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[style=ieee]{biblatex}
\addbibresource{paper.bib}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage[linkcolor = blue]{hyperref}
\newcommand{\MYhref}[3][blue]{\href{#2}{\color{#1}{#3}}}
\textheight = 250mm
% correct bad hyphenation herehttps://www.overleaf.com/project/602635b948e4260c4d50d582
\hyphenation{op-tical net-works semi-conduc-tor}
\newcommand{\todo}[1]{{\color{olive} TODO: #1}}
\pagestyle{fancy}
\fancypagestyle{default}{
    \fancyhf{}
    \renewcommand{\headrulewidth}{0pt}
    \renewcommand{\footrulewidth}{0pt}
    \rfoot{7.\thepage}
}
% Regler til skrivning: 
% 1. Altid nutid
% 2. Brug 1. persons flertal stedord (We/us/our/ours)
% 3. Skriv formelt og uden sammentrækninger (don't do this) 
% 
\pagestyle{fancy}
\begin{document}
\pagestyle{default}
\title{Telemetry Module for the DTU Roadrunners Solar Car}

\author{\IEEEauthorblockN{Victor Alexander Hansen s194027, Steffan Martin Kunoy s194006, Tjark Petersen, s186083}
\IEEEauthorblockA{\textit{Department of Electrical Engineering} \\
\textit{Technical University of Denmark}\\
Kgs. Lyngby, Denmark \\
31015 Introductory project - electrotechnology\\
s194027@student.dtu.dk, s194006@student.dtu.dk, s186083@student.dtu.dk}}
%\author{Victor Alexander Hansen, Steffan Martin Kunoy, Tjark Petersen}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This paper presents the telemetry project for the DTU ROAST solar car. The software and hardware developed for the telemetry system provides a two-way communication between the solar car and a support vehicle. The solar car module can read data from the CAN bus, store it locally in a black box and stream it to the support vehicle, and in return the support vehicle can send commands to the solar car. All messages are secured with an RSA encryption. A graphical user interface enables the support vehicle to stream the CAN data and send commands. In addition, support for a simple UDP network stream was also implemented which can for instance be accessed in Matlab.

@@

\section{Design}
The overall system consists of three separate modules. The Solar Car Unit (SCU) is connected to the CAN bus and communicates with the Support Vehicle Unit (SVU) via a wireless RF connection. The SVU in turn interacts with a laptop in the Support Vehicle through a serial connection. The TelemetryUI program running on the laptop broadcasts the CAN bus message stream on the local network. A block diagram for the system is outlined in Fig. \ref{fig:schematic}. 
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/SystemSchematic.pdf}
    \caption{Overview of the system's components in block diagram form.}
    \label{fig:schematic}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/detailed-can-data-frame-architecture.jpg}
    \caption{Structure of a CAN frame, non-extended.}
    \label{fig:CANframe}
\end{figure}

Fig. \ref{fig:CANframe} shows a detailed structure of the CAN frame. The fields of importance for our project is the message identifier, RTR, control field and data field. The rest of the fields serve as error checking and coordination of communication on the bus.

Each ECU has a unique ID, which the message identifier uses to determine where the CAN message should go to. When two messages try to access the bus, the one with the highest priority (lowest ID) will be granted access to the bus, while the other message have to wait. After the ID comes the RTR bit which determines whether the frame remotely requests or actually sends data to a ECU. The control field, in other literature referred to as "Data Length Code" (DLC), holds the amount of bytes used in the following frame field containing the actual data\cite{DS_ISO}\cite{css}.

@@

\subsection{RF network} % Steffan
Two radio-frequency (RF) transceivers are required to provide a wireless connection between the SCU and SVU. There are many options for radio modules supporting long range communication at the expense of achievable data rate. We chose a trade-off between both by selecting the nRF24L01+ operating on the 2.4 GHz ISM band. The module features a detachable antenna, power amplifier and low-noise amplifier to provide a theoretical range in excess of 1000 m and data transfer rates of up to 2 Mbit/s.

An advantage of using the off-the-shelf RF module is the automated message protocol, known as Enhanced Shockburst$^\copyright$ \cite{shockburst}. Messages are transmitted in packages, structured in flexible byte segments, as shown in Fig. \ref{fig:shockburst}. Only the data payload and address needs to be specified prior to sending a package; the preamble, packet and CRC segments are generated automatically by the RF module.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/EnhancedShockburst.pdf}
    \caption{The Enhanced Shockburst$^\copyright$ packet structure.}
    \label{fig:shockburst}
\end{figure}

@@

The graphical user interface (GUI) included in the telemetry system serves a proof-of-concept purpose. Therefore, scala-swing \cite{scala-swing} was chosen as a framework, due to the effectiveness of the scala \cite{scala} language and the high level of abstraction used when describing a user interface in scala-swing. Furthermore, a GUI implementation in scala can run on any system with a java virtual machine and is therefore cross-platform compatible.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/MessageTypes.pdf}
    \caption{The different message types used in the telemetry system.}
    \label{fig:messageTypes}
\end{figure}

The system is built around \textit{commands} and \textit{message types}. All message types begin with a command byte. The remaining 15 bytes are interpreted based on the message type. As an example, a \texttt{CanTelemetryMsg} message type is shown in Fig. \ref{fig:messageTypes}.

@@

As an extra feature not included in our original specification, we have made an encryption protocol for our telemetry module. The feature can be activated by transmitting an \texttt{ENABLE\_ENCRYPTION} command to the SCU. The encryption is based on the RSA algorithm \cite{rsa}, which essentially generates an encryption key based on the modulus and totient function of two prime numbers. The pairings of prime numbers suited for the encryption protocol can be seen in a table uploaded to DTU ROAST BitBucket \cite{prime}.

@@

A serial connection of up to $\SI{6}{Mbit/s}$ can be achieved using the Teensy 4.0. When sending from the SVU to the TelemetryUI, the message types are converted into JSON strings (see Listing \ref{lst:json}), which are worst case $\SI{117}{bytes}$ long, giving a throughput of $\SI{6410}{s^{-1}}$ messages. The JSON string can be directly used to set the state of an object inside scala which then can be used for further processing.

@@

The SCU is built around the Teensy 3.6 and includes the aforementioned CAN transceiver and RF module. In the solar car, all modules receive power through a molex connector at $\SI{12}{V}$. The molex connector also includes the differential pair of wires of the CAN bus.

@@

A real time clock (RTC) is included on the Teensy 3.6 which can keep the time while the Teensy is powered off by connecting a coin cell battery to the its $V_{bat}$ terminal. This ensures that the generated time stamps associated with a received CAN message are always valid, provided the RTC is synchronized.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/SVU_CircuitDiagram.pdf}
    \caption{The circuit diagram for the SVU.}
    \label{fig:SVU_circuit}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[height=4.5cm]{documentation/images/SVU_threads.pdf}
    \caption{The execution flow of the 2 threads running in the SVU.}
    \label{fig:SVU_threads}
\end{figure}
% threads
\subsection{TelemetryUI}

@@

In the background of the program, multithreading is used to catch new messages received via the serial connection to the SVU, which then are broadcasted to UI elements and a separate thread streaming data to the local network via a UDP port.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{documentation/images/swing.png}
    \caption{A screenshot of the GUI.}
    \label{fig:gui_screenshot}
\end{figure}

@@

\subsection{Throughput limitation}
When testing the throughput capabilities of the system, it was found that the library code used to interface with the RF modules took up to $\SI{1200}{\mu s}$ to send one message, taking around $\SI{200000}{}$ clock cycles.

@@

This shows that there is a lot of performance potential to be gained in optimizing the library code by narrowing it down to only include functionality needed for our use case. Additionally, the message protocol used internally can be optimized to reduce the number of bits sent per message, at the cost of increasing the processing workload.

@@

\section{Conclusion}
The project set out to create a telemetry module for the ROAST solar car and support car, and is successful in most of the points listed in the problem statement. The solar car module can read and write data or commands to and from the CAN bus. The solution for the processing of data in the support car uses both a GUI and streaming of data directly to Matlab. The issue of storing the data locally was solved using a SD card in the Teensy 3.6 as a black box. In addition, a feature for securely encrypting and decrypting the message was implemented.

@@

\section*{Acknowledgment}
The authors would like to thank Martin Schoeberl and Christian Kampp Kruse for their mentoring and assistance throughout the project. The authors would also like to thank Claus Suldrup Nielsen from the department of Mechanical Engineering at DTU.

>>>>

\usepackage[pdftex]{graphicx}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{float}
\usepackage{comment}
\usepackage{textcomp}
\usepackage{siunitx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage[style=ieee]{biblatex}
\addbibresource{paper.bib}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage[linkcolor = blue]{hyperref}
\newcommand{\MYhref}[3][blue]{\href{#2}{\color{#1}{#3}}}
\textheight = 250mm
% correct bad hyphenation herehttps://www.overleaf.com/project/602635b948e4260c4d50d582
\hyphenation{op-tical net-works semi-conduc-tor}
\newcommand{\todo}[1]{{\color{olive} TODO: #1}}
\pagestyle{fancy}
\fancypagestyle{default}{
    \fancyhf{}
    \renewcommand{\headrulewidth}{0pt}
    \renewcommand{\footrulewidth}{0pt}
    \rfoot{7.\thepage}
}
% Regler til skrivning: 
% 1. Altid nutid
% 2. Brug 1. persons flertal stedord (We/us/our/ours)
% 3. Skriv formelt og uden sammentrækninger (don't do this) 
% 
%\pagestyle{fancy}
\begin{document}
\pagestyle{default}
\title{Telemetry Module for DTU Roadrunners' Solar Car}

\author{\IEEEauthorblockN{Victor Alexander Hansen s194027, Steffan Martin Kunoy s194006, Tjark Petersen, s186083}
\IEEEauthorblockA{\textit{Department of Electrical Engineering} \\
\textit{Technical University of Denmark}\\
Kgs. Lyngby, Denmark \\
31015 Introductory Project - Electrotechnology (Group 7)\\
s194027@student.dtu.dk, s194006@student.dtu.dk, s186083@student.dtu.dk}}

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
This paper presents the telemetry project for the DTU ROAST solar car. The software and hardware developed for the telemetry system provides a two-way communication between the solar car and a support vehicle. The solar car module can read data from the CAN bus, store it locally in a black box and stream it to the support vehicle, and in return the support vehicle can send commands to the solar car. All messages are secured with an RSA encryption. A graphical user interface enables the user to stream the CAN data and send commands. In addition, support for a simple UDP network stream was also implemented which can for instance be accessed in Matlab.

@@

\section{Design}
The overall system consists of three separate modules. The Solar Car Unit (SCU) is connected to the CAN bus and communicates with the Support Vehicle Unit (SVU) via a wireless RF connection. The SVU in turn interacts with a laptop in the Support Vehicle through a serial connection. The TelemetryUI program running on the laptop broadcasts the CAN bus message stream on the local network. A block diagram for the system is outlined in Fig. \ref{fig:schematic}. 
\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/SystemSchematic.pdf}
    \caption{Overview of the system's components in block diagram form.}
    \label{fig:schematic}
\end{figure}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/detailed-can-data-frame-architecture.jpg}
    \caption{Structure of a CAN frame, non-extended.}
    \label{fig:CANframe}
\end{figure}

Fig. \ref{fig:CANframe} shows a detailed structure of the CAN frame. The fields of importance for our project is the message identifier, RTR, control field and data field. The rest of the fields serve error checking and coordination of communication on the bus.

Each ECU has a unique ID, which the message identifier uses to determine where the CAN message should go to. When two messages try to access the bus, the one with the highest priority (lowest ID) will be granted access to the bus, while the other message has to wait. After the ID comes the RTR bit which determines whether the frame remotely requests or actually sends data to a ECU. The control field, in other literature referred to as "Data Length Code" (DLC), holds the amount of bytes used in the following frame field containing the actual data\cite{DS_ISO}\cite{css}.

@@

\subsection{RF network} % Steffan
Two RF transceivers are required to provide a wireless connection between the SCU and SVU. There are many options for radio modules supporting long range communication at the expense of achievable data rate. We chose a trade-off between both by selecting the nRF24L01+ operating on the 2.4 GHz ISM band. The module features a detachable antenna, power amplifier and low-noise amplifier to provide a theoretical range in excess of 1000 m and data transfer rates of up to 2 Mbit/s.

An advantage of using the off-the-shelf RF module is the automated message protocol, known as Enhanced Shockburst$^\copyright$ \cite{shockburst}. Messages are transmitted in packages, structured in flexible byte segments, as shown in Fig. \ref{fig:shockburst}. Only the data payload and address needs to be specified prior to sending a package; the preamble, packet and cyclic redundancy check (CRC) segments are generated automatically by the RF module.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/EnhancedShockburst.pdf}
    \caption{The Enhanced Shockburst$^\copyright$ packet structure.}
    \label{fig:shockburst}
\end{figure}

@@

The graphical user interface (GUI) included in the telemetry system serves a proof-of-concept purpose. Therefore, scala-swing \cite{scala-swing} was chosen as a framework, due to the effectiveness of the scala language \cite{scala} and the high level of abstraction used when describing a user interface in scala-swing. Furthermore, a GUI implementation in scala can run on any system with a java virtual machine and is therefore cross-platform compatible.

@@

\begin{figure}[b]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/MessageTypes.pdf}
    \caption{The different message types used in the telemetry system.}
    \label{fig:messageTypes}
\end{figure}

The system is built around \textit{commands} and \textit{message types}. All message types begin with a command byte. The remaining 15 bytes are interpreted based on the message type. All message types can be worked with in an uninterpreted state as a \texttt{BaseTelemetryMsg}. As an example, a \texttt{CanTelemetryMsg} message type is shown in Fig. \ref{fig:messageTypes}.

@@

As an extra feature not included in our original specification, we have made an encryption protocol for our telemetry module. The feature can be activated by transmitting an \texttt{ENABLE\_ENCRYPTION} command to the SCU. The encryption is based on the RSA algorithm \cite{rsa}, which essentially generates an encryption key based on the modulus and totient function of two prime numbers. The pairings of prime numbers suited for the encryption protocol can be seen in a table uploaded to the DTU ROAST BitBucket \cite{prime}.

@@

A serial connection of up to $\SI{6}{Mbit/s}$ can be achieved using the Teensy 4.0. When sending from the SVU to the TelemetryUI, the message types are converted into JSON \cite{json} strings (see Listing \ref{lst:json}), which are worst case $\SI{117}{bytes}$ long, giving a throughput of $\SI{6410}{s^{-1}}$ messages. The JSON string can be directly used to set the state of an object inside scala which then in turn can be used for further processing.

@@

The SCU is built around the Teensy 3.6 and includes the aforementioned CAN transceiver and RF module. In the solar car, all modules receive power through a 4-pin molex connector at $\SI{12}{V}$. The molex connector also includes the differential pair of wires of the CAN bus.

@@

A real time clock (RTC) is included on the Teensy 3.6 which can keep the time while the Teensy is powered off when a coin cell battery is connected to its $V_{bat}$ terminal. This ensures that the generated time stamps associated with a received CAN message are always valid, provided the RTC is synchronized.

@@

\begin{figure}[b]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/SVU_CircuitDiagram.pdf}
    \caption{The circuit diagram for the SVU.}
    \label{fig:SVU_circuit}
\end{figure}
\begin{figure}[t]
    \centering
    \includegraphics[height=4.5cm]{documentation/images/SVU_threads.pdf}
    \caption{The execution flow of the 2 threads running in the SVU.}
    \label{fig:SVU_threads}
\end{figure}

\subsection{TelemetryUI}

@@

In the background of the program, multithreading is used to catch new messages received via the serial connection to the SVU, which then are broadcasted to UI elements and spawn a separate thread streaming data to the local network via a UDP port.

\begin{figure}[b]
    \centering
    \includegraphics[width=\linewidth]{documentation/images/swing.png}
    \caption{A screenshot of the GUI.}
    \label{fig:gui_screenshot}
\end{figure}

@@

\subsection{Throughput limitation}
When testing the throughput capabilities of the system, it was found that the library code used to interface with the RF modules took up to $\SI{1200}{\mu s}$ to send one message, taking around $\SI{200000}{}$ clock cycles on the Teensy 3.6.

@@

This shows that there is a lot of performance potential to be gained in optimizing the library code by narrowing down its functionality and directly working with interrupts. Additionally, the message protocol used internally can be optimized to reduce the number of bits sent per message, at the cost of increasing the processing workload.

@@

\section{Conclusion}
The project set out to create a telemetry module for the ROAST solar car, and is successful in most of the points listed in the problem statement. The solar car module can read and write data or commands to and from the CAN bus. The solution for the processing of data in the support vehicle uses both a GUI and streaming of data directly to Matlab. The issue of storing the data locally was solved using a SD card in the Teensy 3.6 as a black box. In addition, a feature for securely encrypting and decrypting the message was implemented.

@@

\section*{Acknowledgments}
The authors would like to thank Martin Schoeberl and Christian Kampp Kruse for their mentoring and assistance throughout the project. The authors would also like to thank Claus Suldrup Nielsen and the rest of the DTU Roadrunners team for their outstanding cooperation throughout the project.
